---
title: "Introduction to Data Visualization with R"
author: "Scott McCain"
output: html_document
---

Today we will go into some basic introductions to R, but mostly to a data visualization package ggplot2. This is also a first-run at the R notebook, the format that you see here. It allows a mix of plots, code, and text. This is in constrast with an R script by itself, which is just code (hopefully also with comments!)

Data visualization in R can be done in two main avenues: ggplot2, and "base" R graphics. I will demonstrate the use of ggplot2, but base R graphics can be very useful as well. I would argue that ggplot2 is significantly better at exploratory plots, and base R graphics are slightly better for publication level plots (but this is just my humble opinion!). Base R gives you ultimate control over every aspect of your plots.

We will use a dataset that is automatically uploaded to R known as iris. So let's take a look at this dataset!

```{r}
#First we need to upload all the necessary packages into the working space.
#You'll notice that these need to be downloaded first, eg:
#install.packages("ggplot2")
library(ggplot2)
library(magrittr)
```
These are some useful functions for when you've got your data uploaded into R. Uploading data into R can be a challenge by itself. I tend to use mostly ".csv" files, and the `read.csv()` function. We can cover that on another time!
```{r}
#What does top of the dataset look like?
head(iris)#Check out the function tail() also!

#What is the structure? (str function)
str(iris)

#What are the dimensions?
dim(iris)

```
I don't like writing capital letters, so lets change these names...
```{r}
names(iris) <- tolower(names(iris))

```

There are many ways to do the same thing in R. Do you find the above code statement hard to read? I do! That's because we sort of have to read it inside out - that can get very confusing when you have a lot of functions in one line! Here is a neater alternative:
```{r}
names(iris) <- iris %>% names() %>% tolower()
```
We are using the pipe command, ` %>% `. You can read this as "then". So we take the dataframe "iris" *then* get the names, *then* converting to lower case. 



###"Plot the data, plot the data, plot the data" - R.A. Myers

What does it __look__ like!

Start off with the base function `ggplot()`. What are the key arguments? You tell me! (`?ggplot`)

```{r}
names(iris)
ggplot(data = iris, aes(x = sepal.length, y = sepal.width)) + 
  geom_point()
```

This is a pretty great plot already! But I want to colour code by species.
```{r}
ggplot(data = iris, aes(x = sepal.width, y = sepal.length)) + 
  geom_point(aes(colour = species))
```
Great! But I don't want to rewrite that code chunk every time. Let's assign this plot to an object:

```{r}
plot_1 <- ggplot(data = iris, aes(x = sepal.width, y = sepal.length)) + 
                geom_point(aes(colour = species))
print(plot_1)
plot_1 + geom_point(aes(colour = species), size = 3)
```

Note what happens when I put the `size` argument outside of `aes()`. It applies to all of the observations, it's not mapped to a variable. 

What if we want to look at the distribution of these by species?
```{r}
plot_2 <- ggplot(iris, aes(species, sepal.width)) + geom_boxplot() 
print(plot_2)
plot_2 + theme_bw() + geom_jitter()

ggplot(iris, aes(species, sepal.width)) + geom_violin()

```

What if we want to look at the relationship between two variables, but separate it by each species?
```{r}
ggplot(data = iris, mapping = aes(sepal.width, sepal.length)) + 
  geom_point() + 
  facet_grid(~species)
```

This is just scraping the surface of ggplot2, you go from idea to visualization with just a few lines of code. It's extremely effective for exploratory data visualization. To get to a publication level plot, it's debatable which is better: ggplot2 or base graphics. 

There many options for visualizing your data with ggplot2. [This](http://docs.ggplot2.org/current/) website is amazing for exploring the different types!
